---
title: "Project_2"
author: "Sanjeev's SupeRstars: Aaron Zelmanov, Muhammad Hafizudeen Mohamad Saman, Nakul Chadha, Kendall Cohen, Michael Geraci"
date: "2/21/2021"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

data <- read.csv("ProjectA_Listings2013.csv")
library("ggplot2")
library("stats")
library("dplyr")
library("MASS")
library("gmodels")
library("rcompanion")

str(data)
summary(data) 

#factorize 
data$loan_status <- as.factor(data$loan_status)
data$total_open_revolving_accounts <- as.factor(data$loan_status_description)
data$prosper_score <- as.factor(data$prosper_score)
data$income_range <- as.factor(data$income_range)
data$income_range_description <- as.factor(data$income_range_description)
data$income_verifiable <- as.factor(data$income_verifiable)
data$employment_status_description <- as.factor(data$employment_status_description)
data$occupation <- as.factor(data$occupation)
data$borrower_state <- as.factor(data$borrower_state)
data$lender_indicator <- as.factor(data$lender_indicator)
data$is_homeowner <- as.factor(data$is_homeowner)

data$default <- ifelse(data$loan_status == 3, 1, 0)
#remove observations with NA's
any(is.na(data$months_employed))
any(is.na(data$installment_balance))
data <- data[complete.cases(data), ]
```

## {.tabset}
### Introduction 
**Context:** We are working with the ProjectA_Listings2013.csv dataset to better understand loans, interest rates, and defaults. This dataset includes information about: rates, balances, customer types, payments , defaults, etc.

**Audience:** Investors who are interested in taking advantage of arbitrage opportunities 

**Key Question:** Are there elements that the market does not consider a risk factor - but it actually is? Looking in the opposite direction - are there elements that market considers a risk factor but actually it isn’t?


### Basic Exploration 

The goal of this analysis is to identify arbitrage opportunities. We want to look if there are elements that the market does not consider a risk factor - but they actually are, or if there are elements that market considers a risk factor but actually they aren’t. This way, we can construct a way to build portfolios that have positive returns with zero net investment by taking advantage of the inefficiencies in the market.

But first, lets start by looking at the data as a whole. The two most important responses that we wanted to look at are the interest rates and the loan status - whether the borrower defaults or not. Correlations between the two responses and other variables can be helpful in determining which variable is significant in determining the interest rate or whether a borrower defaults. 

```{r, warning = FALSE}
correlationsRates <- as.data.frame(cor(data[sapply(data, function(x) is.numeric(x))])[4,])
correlationsRates %>% ggplot(aes(x=reorder(rownames(correlationsRates), correlationsRates[,1]), y=correlationsRates[,1])) + 
  geom_bar(stat = 'identity') + ggtitle("Correlations with Interest Rate") + labs(x="Variables", y="Correlation") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90))


correlationsDefault <- as.data.frame(cor(data[sapply(data, function(x) is.numeric(x))])[,35])
correlationsDefault %>% ggplot(aes(x=reorder(rownames(correlationsDefault)[], correlationsDefault[,1]), y=correlationsDefault[,1])) + 
  geom_bar(stat = 'identity') + ggtitle("Correlations with Default") + labs(x="Variables", y="Correlation") + theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90))
```

The graphs above show the correlation between the two responses of interest against other variables in the dataset. The leftmost variables are the variables that move in the opposite direction of the variables of interest, ordered in decreasing magnitude, while the rightmost variables are the ones that move in the same direction as the variables of interest, ordered in increasing magnitude. With this graph, we get to see which variables might be good in interpreting the responses. 

We have to note that the magnitude of the correlation is what matters in predicting which variables have a significant impact on the response variables. Negative correlations with high magnitude is likely to produce more predictive power than variables with almost zero correlation with the response.

<br />
<br />
Now, lets take a further look into the variables potentially have a significant effect on the responses.

```{r}
ggplot(data=data, aes(x=amount_funded, y = borrower_rate, colour=borrower_rate)) + geom_jitter(alpha = 0.25) + xlab("Amount Funded") + ylab("Borrower Rate") + ggtitle("Borrower Rate vs Amount Funded") + theme(plot.title = element_text(hjust = 0.5))
```

The data shows that as the amount funded increases, the borrower rate decreases.

```{r, warning = FALSE}
ggplot(data=data, aes(x=as.logical(default), y = borrower_rate, colour=borrower_rate)) + geom_boxplot() + xlab("Defaulted") + ylab("Borrower Rate") + ggtitle("Borrower Rate vs Default Status") + theme(plot.title = element_text(hjust = 0.5))
```

The data above shows that people who defaulted received higher borrowing rate. This is due to the fact that the lender charged the borrower for the credit risk that they took. The more likely the person is to default, the more that the borrower needs to pay so that in case the borrower defaults, the extra rate that the lender charged can cover a portion of the loss from the default.

```{r}
baseplot <- ggplot(data = data, aes(x=bankcard_utilization, y = borrower_rate, colour=borrower_rate)) + xlab("Bankcard Utilization") + ylab ("Borrower Rate") + ggtitle("Borrower Rate vs Bankcard Utilization") + theme(plot.title = element_text(hjust = 0.5))
baseplot+geom_smooth()
```

In the graph above, we can see that the borrower rate increases as the bank card utilization increases. This is most likely due to reason that people who utilize bank card a lot are seen as financially distressed and more likely to default. therefore the credit risk is higher on these people. However, we can also see that when the bank card utilization is fairly low, the borrower rate actually decreases as the utilization increases. This might be because the bank encourages some level of utilization of the card by offering incentives to lower the borrowing rate to customers who utilize credit card until a certain point.

### Linear Regression 
```{r}
set.seed(123)

test_set <- sample(1:nrow(data), 0.2 * nrow(data))

data_train <- data[-test_set, ]
data_test <- data[test_set, ]

data_train_labels <- data[-test_set, "borrower_rate"]
data_test_labels <- data[test_set, "borrower_rate"]
```

```{r}
lm.model <- lm(borrower_rate ~ principal_balance + amount_funded + listing_term + listing_monthly_payment + listing_category_id +  current_delinquencies + public_records_last10_years + inquiries_last6_months + bankcard_utilization + revolving_balance + revolving_available_percent + total_inquiries + delinquencies_over30_days + is_homeowner + loan_status + prosper_rating, data = data_train)
summary(lm.model)

plot(lm.model)

accuracy(list(lm.model), plotit = TRUE, digits = 3)
#summary(step.model)
```

```{r}
#predicted <- ifelse(predict(lm.model, newdata = data_test, type = "response") > 0.5, 1, 0)

#CrossTable(x = data_test_labels, predicted, prop.chisq = FALSE)
```


### Logistic Regression 
```{r}
data_train_labels_2 <- data[-test_set, "default"]
data_test_labels_2 <- data[test_set, "default"]
```

```{r}
logisticM1 <- glm (default ~ number_of_days + principal_balance + amount_funded + borrower_rate + listing_term + listing_monthly_payment + prosper_score + listing_category_id + income_range + stated_monthly_income + income_verifiable + dti_wprosper_loan + months_employed + monthly_debt + current_delinquencies + delinquencies_last7_years + public_records_last10_years + public_records_last12_months + credit_lines_last7_years + inquiries_last6_months + now_delinquent_derog + total_trade_items + amount_delinquent + current_credit_lines + open_credit_lines + bankcard_utilization + installment_balance + real_estate_balance + revolving_balance + real_estate_payment + revolving_available_percent + total_inquiries + satisfactory_accounts + was_delinquent_derog + delinquencies_over30_days + delinquencies_over60_days + delinquencies_over90_days + is_homeowner, data = data, family = "binomial") #all valid variables 
summary(logisticM1)

#logisticM2 <- stepAIC(logisticM1, direction = "both", trace = FALSE)
#summary(logisticM2)

logisticM3 <- glm(default ~ number_of_days + principal_balance + borrower_rate + listing_term + stated_monthly_income + total_inquiries + was_delinquent_derog + principal_balance * listing_term + principal_balance * stated_monthly_income + principal_balance * total_inquiries + amount_funded * stated_monthly_income + I(number_of_days^2) + I(principal_balance^2) + I(amount_funded^2), data = data, family = "binomial")
summary(logisticM3)

test_pred_2 <- predict(logisticM3, newdata = data_test, type = "response")
predicted_results <- ifelse(test_pred_2 > 0.4, 1, 0) 
CrossTable(x = data_test_labels_2, y = predicted_results, prop.chisq=FALSE)
```





### Comparison 




### Conclusion 
